How to build an homogeneous language for heterogeneous platforms

### 2013-05-21 papers/gpce2013/gpce2013.tex
i++

\title{Title Text}
\subtitle{You don’t have to trade abstraction for control}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintainance effort: once you decided to implement a feature on
client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
Even worse, logic parts that need to run on both client-side and server-side have to be duplicated. For instance,
HTML fragments may be built from the server-side when a page is requested by a client, but they may also be built
from the client-side to perform an incremental update subsequent to an user action. How could developers write HTML
fragment definitions once and render them on both client-side and server-side? The more interactive the application
is, the more logic needs to be duplicated between the server-side and the client-side (explain why?).

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
action language natively supported by most of Web clients -- can be used on server-side, and an increasing
number of programming languages or compiler backends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

However, this engineering comfort may come at the price of an inefficient runtime: abstracting over platform
differences often means restricting to a subset of common features and losing opportunities to perform platform
specific optimizations. Performance is a primary concern in Web applications, because they are expected to run on a
broad range of devices, from the powerful desktop personal computer to the less powerful smartphone. “Every 100~ms
delay costs 1\% of sales”, said Amazon in 2006. For instance, because the boundaries of the code sent to the client
are less visible when you share code between client-side and server-side, transitive dependencies may pull a lot of
code on the client, causing a high download overhead. Moreover, generating efficient code for heterogeneous platforms
is hard to achieve in an extensible way: the translation of common abstractions like collections into their native
counterpart (JavaScript arrays on client-side and standard library’s collections on server-side) may be hardcoded in
the compiler, but that would not scale to handle all the abstractions a complete application may use (\eg HTML
fragment definitions, form validation rules, or even some business data type that may be represented differently for
performance reasons).

@@

@@

One way consists in hard-coding, in the language’s compiler, the transformation to each target platform. Figure
\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
real programs. So they need to be at least general purpose languages. It means that for every problem you have to
rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
approach does not scale.

@@

Figure \ref{dedsl} shows a last approach that tries to get the benifits of the two previous approaches, without
their shortcomings.

@@

\begin{itemize}
\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
returning a DOM fragment~;
\item the type system typechecks template definitions and template calls~;
\item the Scala language itself is the expression language~;
\item compiling a template is the same as compiling user code.
\end{itemize}

>>>>

\title{How to build an homogeneous language for heterogeneous platforms}
\subtitle{You don’t have to trade abstraction for control}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: if a piece of logic was implemented on
client-side and it turns out that it needs to be implemented on server-side instead, the code can not be reused and
the feature needs to be completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both
client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
when a page is requested by a client, but they may also be built from the client-side to perform an incremental
update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
both client-side and server-side? The more interactive the application is, the more logic needs to be duplicated
between the server-side and the client-side (explain why?).

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
action language natively supported by most of Web clients -- can be used on server-side, and an increasing
number of programming languages or compiler back-ends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

However, using the same programming language is not enough because the client and server programming environments
are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make it leverage
the native APIs, when possible, for performance reasons?

Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
sales”, said Amazon in 2006.

For instance, because the boundaries of the code sent to the client are less visible when you share code between
client-side and server-side, transitive dependencies may pull a lot of code on the client, causing a high download
overhead. Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way:
the translation of common abstractions like collections into their native counterpart (JavaScript arrays on
client-side and standard library's collections on server-side) may be hard coded in the compiler, but that would not
scale to handle all the abstractions a complete application may use (\eg HTML fragment definitions, form validation
rules, or even some business data type that may be represented differently for performance reasons).

@@

(More on the results metrics)

@@

One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
real programs. So they need to be at least general purpose languages. It means that for every problem you have to
rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
approach does not scale.

@@

Figure \ref{dedsl} shows a last approach that tries to get the benefits of the previous two approaches, without
their shortcomings.

@@

\begin{itemize}
\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
returning a DOM fragment~;
\item the type system type-checks template definitions and template calls~;
\item the Scala language itself is the expression language~;
\item compiling a template is the same as compiling user code.
\end{itemize}
